Type-Safe Modular Parsing

### 2016-12-12 paper/sections/Overview.tex
Rewriting 2.3

It is straightforward to model ASTs by inheritance and write corresponding parsers for each case.

@@

\begin{lstlisting}
trait ExtParser extends BaseParser {
  lexical.delimiters += ("\\", ".")
  val pExtExpr: PackratParser[Expr] = pExpr ||| pLam ||| "(" ~> pExtExpr <~ ")"
  val pLam: PackratParser[Expr] =
    "\\" ~> ident ~ ("." ~> pExtExpr) ^^ { case x ~ e => new Lam(x, e) }
}
\end{lstlisting}

@@

\paragraph{Open Parsing}
The example in Section \ref{} can be rewritten as follows:

\begin{lstlisting}
type Open[T] = (=> T) => T

// Expr, Var and App are the same as previous section
trait ExprParser {
  val pVar: Open[PackratParser[Expr]] =
    _ => ident ^^ { x => new Var(x) }
  val pApp: Open[PackratParser[Expr]] =
    self => self ~ self ^^ { case e1 ~ e2 => new App(e1, e2) }
}
\end{lstlisting}

@@

\lstinline{pApp} takes a parser of
\lstinline{Expr} as an argument, and uses it to parse two consecutive
expressions as an application. This argument \lstinline{e} represents
the parser of \emph{the whole expression}. In this particular case, it
is in fact the combination of \lstinline{pVar} and
\lstinline{pApp}. We can replace it by \lstinline{pVar | pApp} and the
code still works. However, using an abstract argument provides
extensibility to the parser. Because once the abstract syntax is
extended later, the parser of the whole expression will change
correspondingly. Then we have to edit existing code if it is
hard-coded. The parameterization technique used in \lstinline{pApp} is
called \textit{open recursion}, which is discussed in section
\ref{sec:openandparsing}.

\paragraph{Parsing Components}\bruno{consider removing and/or moving
  to later parts of the paper}
We can use a trait to glue them together, and define
\lstinline{pVarApp} as a combination of the two parsers. A type
synonym \lstinline{Fix} is used to shorten the types.

\begin{lstlisting}
type Fix[T] = (=>T) => T

trait ExprParser {
  lexical.delimiters += ("(", ")")

  val pVar: PackratParser[Expr] =
    ident ^^ Var
  val pApp: Fix[PackratParser[Expr]] =
    e => e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
  val pVarApp: Fix[PackratParser[Expr]] =
    e => pApp(e) | pVar | "(" ~> e <~ ")"
}
\end{lstlisting}
\bruno{Fix is not the right name. A better name is Open (for open
  recursion).}
\bruno{I'm not sure that showing the code for language components is
  essential at this point. This can be shown latter in the paper.}

\paragraph{Client Code}
Finally, we are able to parse \lstinline{Expr} frome strings as shown in the code below. The \lstinline{fix} function here is just the standard \textit{fixpoint}, it is explained in section \ref{subsec:openrecursion}.

\begin{lstlisting}
def parse[E](p: Fix[PackratParser[E]])(inp: String): E = {
  val t = phrase(fix(p))(new lexical.Scanner(inp))
  if (t.successful) t.get else scala.sys.error(t.toString)
}

println(parse(new ExprParser {}.pVarApp)("x y"))
\end{lstlisting}

It will print the AST as \lstinline{App(Var(x),Var(y))}.

%\subsection{Extend and Reuse Parsers}\label{subsec:overview-extend}

\paragraph{Extensibility} To demonstrate the advantages of writing
parsers with open recursion, we extend the language to add lambda
abstractions as a new case of expressions. The language then
becomes the famous \textit{untyped lambda calculus}.

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <var>
    \alt <expr> <expr>
    \alt `\\' <var> `.' <expr>
\end{grammar}

With Scala's case classes, extending the abstract syntax \lstinline{Expr} is easy.

\begin{lstlisting}
case class Lam(x: String, e: Expr) extends Expr
\end{lstlisting}

Since we already have the parser for variables and applications, we
want to reuse it and build a new parser by extending it. It is just
what the code below does. Notice that the old code is not touched,
hence separate compilation is obtained.

\begin{lstlisting}
trait LambdaParser extends ExprParser {
  lexical.delimiters += ("\\", ".")

  val pLam: Fix[PackratParser[Expr]] =
    e => "\\" ~> ident ~ ("." ~> e) ^^ { case x ~ b => Lam(x, b) }
  val pVarAppLam: Fix[PackratParser[Expr]] =
    e => pLam(e) | pVarApp(e)
}
\end{lstlisting}

In the new parser \lstinline{LambdaParser} extended from the old one,
we write a new parsing function \lstinline{pLam} for the new case
lambda abstraction. And compose it with \lstinline{pVarApp}, which we
already have in the old parser \lstinline{ExprParser}. Open recursion
is critical to achieve such type-safe compositions, because wherever
we refer to the parser of the whole expression, it is updated to cover
all of the three case we have now.

\begin{lstlisting}
println(parse(new LambdaParser {}.pVarAppLam)("(\\x.x) y"))
\end{lstlisting}

The code above will print \lstinline{App(Lam(x,Var(x)),Var(y))} from parsing \lstinline{"(\\x.x)} \lstinline{y"}.

>>>>

It is straightforward to model ASTs by inheritance and write corresponding parsers for each case. \huang{This is the first time to demonstrate code in this paper. Readers may not know what those combinators/types are. Refer to the library docs here? move the 'scala parsing library' section to the beginning as 'background'?}

@@

\begin{lstlisting}
trait ExtParser extends BaseParser {
  lexical.delimiters += ("\\", ".")
  val pExtExpr: PackratParser[Expr] =
    pVar ||| pApp ||| pLam ||| "(" ~> pExtExpr <~ ")"
  val pLam: PackratParser[Expr] =
    "\\" ~> ident ~ ("." ~> pExtExpr) ^^ { case x ~ e => new Lam(x, e) }
}
\end{lstlisting}

@@

\paragraph{Open Parsing} Consider the expression language of only variables and applications in the previous section. The parser can be rewritten as follows:

\huang{todo: polish it}
\begin{lstlisting}
type Open[T] = (=> T) => T

// Expr, Var and App are the same as previous
trait BaseParser {
  lexical.delimiters += ("(", ")")
  val pVar: Open[PackratParser[Expr]] =
    self => ident ^^ { x => new Var(x) }
  val pApp: Open[PackratParser[Expr]] =
    self => self ~ self ^^ { case e1 ~ e2 => new App(e1, e2) }
  val pExpr: Open[PackratParser[Expr]] =
    self => pVar(self) ||| pApp(self) ||| "(" ~> self <~ ")"
}
\end{lstlisting}

According to the definition of type synonym \lstinline{Open[T]}, every parse function takes a parser of \lstinline{Expr} as a parameter, namely \lstinline{self}. This parameter \lstinline{self} represents \emph{the parser} of expressions, which is dynamiclly explained by the argument. Using an abstract parameter provides extensibility to parsers, because the recursive calls are not hard-coded here --- they are dynamic and can be changed corresponding to the grammar!

We take \lstinline{pApp} as an example again. It uses \lstinline{self} instead of hard-coded \lstinline{pExpr} for recursive parsing. Depending on the real argument, \lstinline{self} can be a parser which covers variables and applications, or a parser which also supports the lambda case.

\paragraph{Client Code} Since the type of parsing functions are changed, we can not apply the parsers directly. The \lstinline{fix} function below, which is the the standard \textit{fixpoint combinator}, is used to `close' the open recursion.

\begin{lstlisting}
def fix[T](f: Open[T]): T = { lazy val a: T = f(a); a }

def parse[E](p: Open[PackratParser[E]])(inp: String): E = {
  val t = phrase(fix(p))(new lexical.Scanner(inp))
  if (t.successful) t.get else scala.sys.error(t.toString)
}

println(parse(new BaseParser {}.pExpr)("x y"))
\end{lstlisting}

@@

\paragraph{Extensibility} As same as what we did in previous section, we extend the grammar by adding lambda abstractions as a new case. Notice that the old code is not touched, hence separate compilation is obtained.

\begin{lstlisting}
trait ExtParser extends BaseParser {
  lexical.delimiters += ("\\", ".")
  val pLam: Open[PackratParser[Expr]] =
    self => "\\" ~> ident ~ ("." ~> self) ^^ { case x ~ b => new Lam(x, b) }
  val pExtExpr: Open[PackratParser[Expr]] =
    self => pLam(self) ||| pExpr(self)
}
\end{lstlisting}

\huang{todo: rewrite}
In the new parser \lstinline{LambdaParser} extended from the old one,
we write a new parsing function \lstinline{pLam} for the new case
lambda abstraction. And compose it with \lstinline{pVarApp}, which we
already have in the old parser \lstinline{ExprParser}. Open recursion
is critical to achieve such type-safe compositions, because wherever
we refer to the parser of the whole expression, it is updated to cover
all of the three case we have now.

\begin{lstlisting}
println(parse(new ExtParser {}.pExtExpr)("(\\x.x) (\\y.y)"))
\end{lstlisting}

Parsing with open recursion will be discussed further in section \ref{sec:openandparsing}.
